Group 10 Final Project

### 2024-08-03 starter-analysis-exercise/products/report/manuscript_final-2a.qmd
Updated statements and in-depth

---
title: "Honey Bee Colonies Impacted by Varroa, American Foulbrood and Global Warming"
subtitle: ""
author: Edward Cruz, Jr., Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
date: today
format:
  html:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../citations.bib 
csl: ../apa.csl
editor: 
  markdown: 
    wrap: 72
---

@@

Group Ten is conducting a comprehensive analysis of historical data from
various public agencies to evaluate the impact of Varroa mites, American
Foulbrood, and global warming on hive losses in the United States.This
research utilizes extensive datasets from the National Agricultural
Statistics Service, the Agricultural Statistics Board, and the United
States Department of Agriculture (USDA), encompassing multiple years of
data. The data highlights the topic on hive losses attributed to mites,
bacterial infections, and environmental factors related to global
warming. Through advanced data visualization techniques in R, we aim to
demonstrate and validate the detrimental effects of these factors on
honeybee colonies, highlighting the consequent implications for honey
production and broader food security. ***Dennis and Kemp's study on
honey bee hive collapse [@hivecollapse] provides important insights into
Allee effects and ecological resilience.*** ***Warmer climate may impact
bees[@2024warmer]***

@@

Honeybees have served as nature's pollinators for centuries. With their
relationship with humans documented as far back as ancient Egyptian and
Hindu cultures. Historically, humans have maintained beehives and
utilized honey for its medicinal properties in various civilizations,
including the Egyptians, Assyrians, Chinese, Greeks, and Romans. Honey's
natural antibacterial qualities made it a valuable treatment for wounds.
This was a practice continued by Romans and Russians during World War I.
Honeybees and other pollinators are essential for food production and
nutritional security. But even though they face numerous survival
challenges Varroa mites pose a significant threat to bee colonies. These
tiny red-brown parasites live on adult honeybees and reproduce on larvae
and pupae. Another major threat is American Foulbrood Disease (AFB).
This is caused by the bacterium Paenibacillus larvae, which is fatal to
honeybee larvae and requires incineration of infected hives to prevent
its spread. The impact of climate change on honeybee colony losses is a
growing area of research. While correlations have been found between
higher winter temperatures and increased colony losses, the effects of
warmer autumn and winter temperatures on colony dynamics and survival
require further investigation. ***Dennis and Kemp's study on honey bee
hive collapse [@hivecollapse] provides important insights into Allee
effects and ecological resilience.*** [@usda_honey_bees;
@usda_nass]\*\*\*

***Warmer climate may impact bees[@2024warmer]***

### Composition of Honey

The first image is a pie chart that illustrates the composition of honey
by percentage. The main components are:

@@

# Plot with ggplot2
ggplot(honey_data, aes(x = "", y = percentage, fill = component)) +
  geom_bar(width = 0.8, stat = "identity", color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = honey_data$color) +
  geom_text(aes(x=1.8, label = paste0(percentage, "%")), position = position_stack(vjust = 0.6), size = 3, color = "black") +
  labs(
    title = "COMPOSITION OF HONEY",
    subtitle = "Illustration of honey components by percentage",
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, family = "lobster"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 12)
  )
```

@@

### Spread of Varroa Mites

The second image is a map illustrating the spread of Varroa mites in the
United States from 1986 to 1995. The map uses different colors to
represent the year of the first appearance of Varroa mites in each
state:

@@

[@varroamap] ***Varroa Map of infestation periods***

@@

[@epa_climate] ***Global warming trends from 1900-2024***

@@

Bee colonies maintained by beekeepers are considered livestock by the
USDA due to their ability to produce honey, a consumable food item, and
their essential role in assisting farmers with pollination crop seasons.
Given the importance of bee colonies in agriculture, it was logical to
source data from the following two authoritative websites: 1. USDA
National Agricultural Statistics Service (NASS): This site provides
comprehensive agricultural data, including statistics on honey
production and colony health. 2. Bee Informed Partnership: This site
offers detailed insights and research on bee colony management and
health, contributing valuable information on the status and trends of
bee populations. [@usda_honey_bees]

Index Catalog // USDA Economics, Statistics and Market Information
System. (n.d.-a).
https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en\
USDA - National Agricultural Statistics Service - Surveys - honey bee
surveys and reports. (n.d.).
https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/

@@

Hypothesis: "The negative impacts of mites, bacterium, and global
warming have detrimental effects on honeybee colonies in the United
States and Texas, which in turn will lead to a decline in honey
production and negatively impact food production." This hypothesis can
be tested and validated through a visualization of outcomes using R,
demonstrating the relationship between these factors and their effects
on honeybee colonies.

@@

![](/images/dead-bees-hive-colony-collapse-600nw-2289939105.webp)
{{< pagebreak >}}

# Methods

*Describe your methods. That should describe the data, the cleaning
processes, and the analysis approaches. You might want to provide a
shorter description here and all the details in the supplement.*

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Install necessary packages
if (!require("DiagrammeR")) install.packages("DiagrammeR")

# Load the DiagrammeR package
library(DiagrammeR)

# Create the flowchart using DiagrammeR
grViz("
  digraph Data_Analysis_Flowchart {
    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 45, labelloc = 't']

    node [style = filled, fontsize = 30, width = 2, height = 2]

    Define_Objectives [label = 'Define Objectives', shape = oval, color = lightblue]
    Collect_Data [label = 'Collect Data', shape = box, color = lightgrey]
    Clean_Data [label = 'Clean Data', shape = ellipse, color = lightyellow]
    Analyze_Data [label = 'Analyze Data', shape = diamond, color = lightcoral]
    Interpret_Results [label = 'Interpret Results', shape = trapezium, color = lightgrey]
    Visualize_Data [label = 'Visualize Data', shape = parallelogram, color = lightpink]
    Compile_Report [label = 'Compile Report', shape = house, color = lightcyan]
    Review_Revise [label = 'Review & Revise', shape = invhouse, color = lightgoldenrodyellow]
    Present_Report [label = 'Present Report', shape = hexagon, color = lightseagreen]

    Define_Objectives -> Collect_Data
    Collect_Data -> Clean_Data
    Clean_Data -> Analyze_Data
    Analyze_Data -> Interpret_Results
    Interpret_Results -> Visualize_Data
    Visualize_Data -> Compile_Report
    Compile_Report -> Review_Revise
    Review_Revise -> Present_Report
  }
")

@@

Data was derived from the United States Department of Agriculture (USDA) website, which permits data sets to be downloaded in CSV(comma-separated values) format.

@@

We decided to clean out our data from a few different datasets. We had
to remove blank spaces and columns that were not pertinent to our
analysis. We then filtered out other observations that did not directly
deal with the data we are exploring. We are looking for cause of death
to bee colonies and how they are affected by mites and climate change so
we wanted to single out data that represented the losses so we can
explore the different states by year and determine how the colonies were
affected.

@@

here()
# Create a 'data' directory within the project
dir.create(here("data"))

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# View data str to see what class they are.

str(data_filtered)
```

### Boxplot of Value by Year and Data Item

The third image is a boxplot comparing the values for "LOSS, COLONY
COLLAPSE DISORDER" and "LOSS, DEADOUT" from 2015 to 2017. The boxplot
provides insights into: - The central tendency and spread of losses due
to both disorders. - Presence of outliers indicates that there are
occasional, significantly higher losses. - Comparative analysis shows
that both types of losses have varying distributions each year.

This indicates that while both types of losses are present each year,
the magnitude and frequency of extreme loss events can differ,
suggesting variability in the factors contributing to colony losses.

@@

str(data_filtered)

@@

### Scatterplot of Value by State ANSI

The fourth image is a scatterplot showing values for different states
(ANSI coded) across the years 2015 to 2017. Key points include: - Each
color represents a different year (2015: Red, 2016: Green, 2017:
Blue). - The scatterplot reveals the distribution of values across
states and highlights any trends or patterns over the three years.

This scatterplot indicates that while some states experience
consistently high or low values, others show significant variability
year over year. This could suggest regional differences in factors
affecting bee colony health.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Scatterplot of Value by State.ANSI
ggplot(data_filtered, aes(x = State.ANSI, y = Value, color = factor(Year))) +
  geom_point() +
  labs(title = "Scatterplot of Value by State.ANSI", x = "State.ANSI", y = "Value") +
  theme_minimal()

@@

### Boxplot of Value by Period and Data Item

The fifth image is a boxplot comparing the values for "LOSS, COLONY
COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
THRU DEC**

Key insights include: - Seasonal trends in colony losses, with some
periods showing higher median losses or more variability. - Outliers
indicate extreme loss events occurring in specific periods.

This seasonal analysis suggests that certain times of the year are more
prone to higher losses, potentially due to environmental factors,
resource availability, or management practices.

@@

### Boxplot of Value by Period and Data Item

The fifth image is a boxplot comparing the values for "LOSS, COLONY
COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
THRU DEC**

Key insights include: - Seasonal trends in colony losses, with some
periods showing higher median losses or more variability. - Outliers
indicate extreme loss events occurring in specific periods.

This seasonal analysis suggests that certain times of the year are more
prone to higher losses, potentially due to environmental factors,
resource availability, or management practices.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Descriptive Statistics
print("Descriptive Statistics:")
summary(data_clean)

@@

![](/images/Bee%20Feeding.jpg){fig-align="center" width="313"}

@@

## Full analysis

##The full analysis is still pending###

# References

>>>>

---
title: "Group 10 Final Project"
output: html_document
---

@@

Group Ten is conducting a comprehensive analysis of historical data from
various public agencies to evaluate the impact of Varroa mites, American
Foulbrood, and global warming on hive losses in the United States.This
research utilizes extensive datasets from the National Agricultural
Statistics Service, the Agricultural Statistics Board, and the United
States Department of Agriculture (USDA), encompassing multiple years of
data. The data highlights the topic on hive losses attributed to mites,
bacterial infections, and environmental factors related to global
warming. Through advanced data visualization techniques in R, we aim to
demonstrate and validate the detrimental effects of these factors on
honeybee colonies, highlighting the consequent implications for honey
production and broader food security. ***Dennis and Kemp's study on
honey bee hive collapse [@hivecollapse] provides important insights into
Allee effects and ecological resilience.***

@@

Honeybees have served as nature's pollinators for centuries. With their
relationship with humans documented as far back as ancient Egyptian and
Hindu cultures. Historically, humans have maintained beehives and
utilized honey for its medicinal properties in various civilizations,
including the Egyptians, Assyrians, Chinese, Greeks, and Romans. Honey's
natural antibacterial qualities made it a valuable treatment for wounds.
This was a practice continued by Romans and Russians during World War I.
Honeybees and other pollinators are essential for food production and
nutritional security. But even though they face numerous survival
challenges Varroa mites pose a significant threat to bee colonies. These
tiny red-brown parasites live on adult honeybees and reproduce on larvae
and pupae. Another major threat is American Foulbrood Disease (AFB).
This is caused by the bacterium Paenibacillus larvae, which is fatal to
honeybee larvae and requires incineration of infected hives to prevent
its spread. The impact of climate change on honeybee colony losses is a
growing area of research. While correlations have been found between
higher winter temperatures and increased colony losses, the effects of
warmer autumn and winter temperatures on colony dynamics and survival
require further investigation. ***Dennis and Kemp's study on honey bee
hive collapse [@hivecollapse] provides important insights into Allee
effects and ecological resilience.***[@usda_honey_bees; @usda_nass]

# Plot with ggplot2
ggplot(honey_data, aes(x = "", y = percentage, fill = component)) +
  geom_bar(width = 0.8, stat = "identity", color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = honey_data$color) +
  geom_text(aes(x=1.8, label = paste0(percentage, "%")), position = position_stack(vjust = 0.6), size = 3, color = "black") +
  labs(
    title = "COMPOSITION OF HONEY",
    subtitle = "Illustration of honey components by percentage",
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, family = "lobster"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 12)
  )
```
### Composition of Honey

Compositions of honey is a pie chart that illustrates the composition of honey
by percentage. The main components are:

@@

@@

```
### Spread of Varroa Mites

the spread of varroa mites image is a map illustrating the spread of Varroa mites in the
United States from 1986 to 1995. The map uses different colors to
represent the year of the first appearance of Varroa mites in each
state:

@@

***[@varroamap] Varroa Map of infestation periods***

@@

The chart provided illustrates the changes in temperature anomalies over the years, from 1900 to approximately 2025. The temperature anomaly, measured in degrees Celsius (°C), is plotted on the y-axis, while the years are plotted on the x-axis. The data is divided into three different measurements: Earth's surface, lower troposphere (RSS), and lower troposphere (UAH), each represented by different colors (blue, red, and green respectively).

### Detailed Description:
- **Trend Analysis**:
  - From 1900 to around 1975, the Earth's surface temperature shows considerable variability with some notable periods of cooling and warming.
  - Post-1975, there is a distinct upward trend in temperature anomalies for all three measurements, indicating a general warming trend.
- **Measurement Comparison**:
  - The Earth's surface (blue line) shows the most extended historical data and fluctuates more in the earlier years compared to the troposphere measurements.
  - The lower troposphere measurements (RSS in red and UAH in green) start around 1975 and generally follow the same upward trend as the surface measurements but with less variability.
  - The RSS and UAH measurements align closely with each other post-2000, indicating consistent trends in the lower troposphere temperature anomalies.
- **Anomalies & Peaks**:
  - The Earth's surface shows significant peaks and troughs, especially in the early 20th century and around the year 2000.
  - The lower troposphere measurements show peaks and align closely, particularly after the year 2000, reflecting similar patterns in temperature anomalies.

### Key Points:
- The overall trend shows an increase in temperature anomalies from 1900 to 2025.
- Significant variability is seen in Earth's surface temperatures in the early years.
- Post-1975, a clear warming trend is observed across all measurements.
- Lower troposphere data (RSS and UAH) begins around 1975 and follows the Earth's surface trend with less variability.
- Post-2000, the RSS and UAH data closely align, indicating consistent temperature trends in the lower troposphere.
- There are noticeable peaks in the temperature anomalies around the year 2000 and later, reflecting periods of significant temperature increases.

### Impact on Bees, Varroa Mites, and Colony Loss:
- **Bees**:
  - **Thermal Stress**: Increased temperature anomalies can cause thermal stress on bees, affecting their ability to forage, navigate, and perform essential colony tasks.
  - **Floral Resources**: Warming temperatures can alter the availability and distribution of floral resources, impacting bee nutrition and health.
  - **Reproduction Cycles**: Changes in temperature can disrupt the reproductive cycles of bees, potentially leading to mismatches with the availability of pollen and nectar.

- **Varroa Mites**:
  - **Increased Reproduction**: Warmer temperatures can accelerate the reproduction rates of varroa mites, exacerbating infestations within bee colonies.
  - **Extended Activity Period**: Higher temperatures can lengthen the active period of varroa mites throughout the year, leading to prolonged periods of stress on bee populations.

- **Colony Loss**:
  - **Health Decline**: Combined effects of thermal stress on bees and increased varroa mite infestations can lead to a decline in colony health.
  - **Colony Collapse Disorder (CCD)**: Temperature anomalies can contribute to conditions that favor Colony Collapse Disorder, where the majority of worker bees disappear, leaving behind the queen and immature bees.
  - **Mortality Rates**: Sustained temperature increases can lead to higher mortality rates within bee colonies, significantly impacting beekeeping and agricultural pollination services.

### Conclusion:
The chart highlights a clear trend of increasing temperature anomalies, which can have profound implications on bee populations, varroa mite dynamics, and overall colony health. Understanding these trends is crucial for developing strategies to mitigate the adverse effects on bees and ensure the sustainability of pollination services essential for ecosystems and agriculture.

[@epa_climate_indicator] ***Global warming trends from 1900-2024***

@@

### Data Sources

Bee colonies maintained by beekeepers are classified as livestock by the USDA due to their production of honey, a consumable food item, and their critical role in pollination during crop seasons. Given the importance of bee colonies in agriculture, we sourced data from two authoritative websites:

1. **USDA National Agricultural Statistics Service (NASS)**:
   - Provides comprehensive agricultural data, including statistics on honey production and colony health.
   - Source: [USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports](https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/).

2. **Bee Informed Partnership**:
   - Offers detailed insights and research on bee colony management and health, contributing valuable information on the status and trends of bee populations.
   - Source: [Index Catalog // USDA Economics, Statistics and Market Information System](https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en).

These sources provided reliable and extensive data necessary for analyzing the factors affecting bee colonies, honey production, and the broader implications for agriculture.

@@

### Hypotheses

**Hypothesis**:
"The negative impacts of mites, bacterium, and global warming have detrimental effects on honeybee colonies in the United States and Texas, which in turn will lead to a decline in honey production and negatively impact food production." This hypothesis can be tested and validated through a visualization of outcomes using R, demonstrating the relationship between these factors and their effects on honeybee colonies.

**Null Hypothesis (H0)**:
The impacts of mites, bacterium, and global warming do not have a significant detrimental effect on honeybee colonies in the United States and Texas, and there is no consequent decline in honey production or negative impact on food production.

**Alternative Hypothesis (H1)**:
The impacts of mites, bacterium, and global warming have a significant detrimental effect on honeybee colonies in the United States and Texas, leading to a decline in honey production and negatively impacting food production.

These hypotheses will be tested using various data visualization techniques in R, allowing us to explore and validate the relationships between these factors and their effects on honeybee colonies.

@@

![](/images/dead-bees-hive-colony-collapse-600nw-2289939105.webp)

#Citation section To cite other work (important everywhere, but likely
happens first in introduction), make sure your references are in the
bibtex file specified in the YAML header above and have the right bibtex
key. Then you can include like this:

Examples of reproducible research projects can for instance be found in
[@mckay2020; @mckay2020a].

#Here are the BibTex entries for citations:

Using this in RMarkdown: #Make sure to save these entries in a file
names'reference.bib' and include it in your R Markdown document like
this:

# Methodology

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(DiagrammeR)

DiagrammeR::grViz("
digraph G {

  node [shape = box, style = filled, color = lightblue, fontname = Helvetica]

  subgraph cluster_0 {
    label = 'Analytical Report Workflow'

    Introduction -> Data_Acquisition -> Data_Cleaning_Preparation -> Exploratory_Data_Analysis -> Data_Visualization -> Advanced_Statistical_Analysis -> Results_Discussion -> Conclusion -> Appendix -> References
  }

  Introduction [label = '1. Introduction']
  Data_Acquisition [label = '2. Data Acquisition']
  Data_Cleaning_Preparation [label = '3. Data Cleaning and Preparation']
  Exploratory_Data_Analysis [label = '4. Exploratory Data Analysis (EDA)']
  Data_Visualization [label = '5. Data Visualization']
  Advanced_Statistical_Analysis [label = '6. Advanced Statistical Analysis']
  Results_Discussion [label = '7. Results and Discussion']
  Conclusion [label = '8. Conclusion']
  Appendix [label = '9. Appendix']
  References [label = '10. References']
}
")
```

#knitr::include_graphics(here("assets","antigen-recognition.png"))
```

@@

### Data Acquisition

We sourced our data from the United States Department of Agriculture (USDA), leveraging their comprehensive and authoritative datasets. This data provided crucial information for our analysis, including statistics on honey production, colony health, and factors affecting bee populations. Utilizing USDA data ensured the reliability and relevance of our analysis, as it encompasses extensive, well-documented records pertinent to agricultural and ecological studies.

@@

### Data Cleaning and Preparation

We integrated data from multiple datasets to ensure a comprehensive analysis. The cleaning process involved several key steps:

1. **Removal of Irrelevant Data**:
   - **Blank Spaces**: We eliminated any blank spaces within the datasets to ensure data integrity.
   - **Non-Pertinent Columns**: Columns that were not directly related to our analysis objectives were removed to streamline the dataset.

2. **Filtering Observations**:
   - **Focus on Relevant Data**: We filtered out observations that did not directly pertain to our study, specifically those not related to the causes of bee colony losses.
   - **Loss Data**: The dataset was refined to single out data representing the losses attributed to mites and climate change, allowing us to focus on these critical factors.

3. **Exploratory Focus**:
   - **State and Year Analysis**: By focusing on data that represented colony losses across different states and years, we aimed to explore the geographical and temporal impacts of these factors on bee colonies.

These steps ensured that our dataset was clean, relevant, and well-prepared for subsequent analysis, enabling a focused exploration of the causes of bee colony losses and their relationship with mites and climate change.

@@

# Create a 'data' directory within the project
dir.create(here("data"))

@@

@@

@@

### Analysis of Boxplot

The boxplot illustrates the distribution of values for "LOSS, COLONY COLLAPSE DISORDER" and "LOSS, DEADOUT" across the years 2015 to 2017. A few key observations can be made from the data:

**Central Tendency and Spread**:
   - The median values for both types of losses (represented by the central line in each box) are relatively low across all years, indicating that the majority of losses are concentrated towards the lower end of the scale.

**Yearly Comparison**:
   - For the year 2015, the losses due to Colony Collapse Disorder and Deadout are higher compared to the subsequent years. This can be observed from the height of the boxes and the spread of the data points.
   - There is a noticeable decrease in the losses for both categories in 2016 and 2017. The boxes for these years are smaller and closer to the x-axis, indicating lower values.

**Outliers**:
   - There are significant outliers present in all years for both types of losses. These outliers indicate that while most of the data points are clustered around the lower values, there are instances of very high losses that deviate from the norm.

**Comparison between Loss Types**:
   - The distribution of "LOSS, DEADOUT" appears to have more variability compared to "LOSS, COLONY COLLAPSE DISORDER", particularly in 2015 and 2016. This is indicated by the wider interquartile ranges and more spread out data points.

**Overall Trends**:
   - Overall, the boxplot suggests a decline in both types of losses from 2015 to 2017. However, the presence of outliers in each year indicates that there are still occasional severe loss events.

This highlights the trends and variations in honeybee colony losses due to Colony Collapse Disorder and Deadout over a three-year period, providing insights into the distribution and severity of these losses.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Scatterplot of Value by State.ANSI
ggplot(data_filtered, aes(x = State.ANSI, y = Value, color = factor(Year))) +
  geom_point() +
  scale_y_continuous(labels = comma) +
  labs(title = "Scatterplot of Value by State.ANSI", x = "State.ANSI", y = "Value") +
  theme_minimal()

@@

### Analysis of Scatterplot

The scatterplot visualizes the distribution of values across different states (represented by State ANSI codes) for the years 2015, 2016, and 2017. Several key observations can be drawn from the data:

**Value Distribution Across States**:
   - The values vary significantly across different states. Most states have relatively low values, but there are notable exceptions with very high values.

**Yearly Trends**:
   - In 2015, there are multiple high values, especially at the lower end of the State ANSI codes. These are represented by the red dots.
   - In 2016, represented by green dots, there are fewer high values, but some extreme outliers are present, such as the one exceeding 200,000.
   - In 2017, indicated by blue dots, the values are generally lower compared to the previous years, with fewer extreme outliers.

**Outliers**:
   - Each year has significant outliers that deviate from the general trend of low values across states. These outliers suggest occasional severe losses in specific states.

**State-Specific Trends**:
   - Certain states (e.g., those with lower ANSI codes) exhibit higher variability and more frequent high values compared to others.
   - States towards the middle and higher end of the ANSI scale generally show more consistent, lower values.

**Comparative Analysis**:
   - The plot indicates that while overall losses may be high in certain states, the distribution and frequency of these high values vary across years. This suggests possible changes in factors affecting bee colony losses over time.

This shows the distribution and variability of bee colony losses across different states over three years, providing insights into state-specific trends and the presence of extreme loss events.

@@

### Analysis of Boxplot

The boxplot illustrates the distribution of values for "LOSS, COLONY COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the year. The data is segmented into four periods: APR-THRU-JUN, JAN-THRU-MAR, JUL-THRU-SEP, and OCT-THRU-DEC. Several key observations can be made from the data:

**Central Tendency and Spread**:
   - The median values for both types of losses are low across all periods, indicating that most losses are concentrated towards the lower end of the scale.
   - The interquartile ranges (IQRs) for "LOSS, DEADOUT" are relatively consistent across periods, suggesting stable variability in losses throughout the year.

**Period Comparison**:
   - The period JUL-THRU-SEP shows the highest median value for "LOSS, DEADOUT," indicating that this period might be particularly challenging for bee colonies.
   - The period JAN-THRU-MAR has a few outliers but generally shows lower losses compared to other periods.

**Outliers**:
   - Significant outliers are present in all periods for both types of losses. These outliers suggest occasional severe loss events that deviate significantly from the norm.
   - The period APR-THRU-JUN shows the highest outlier for "LOSS, COLONY COLLAPSE DISORDER," indicating some extreme events during this time.

**Comparison between Loss Types**:
   - "LOSS, DEADOUT" consistently shows a wider range of values compared to "LOSS, COLONY COLLAPSE DISORDER" across all periods, suggesting more variability in deadout losses.

**Overall Trends**:
   - The plot suggests that while losses occur throughout the year, there are specific periods (e.g., JUL-THRU-SEP) where losses might be higher.
   - The consistent presence of outliers indicates that, despite general trends, extreme loss events are a recurring issue.

This boxplot shows insights into the seasonal variability of bee colony losses due to Colony Collapse Disorder and Deadout, highlighting specific periods with higher losses and the presence of extreme loss events.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Descriptive Statistics
print("Descriptive Statistics:")
summary(data_clean)
str(data_clean)
```

@@

![](/images/Bee%20Feeding.jpg){fig-align="center"}

@@
